Stage-Splitting a Modal Language

### 2014-10-06 paper/intro.tex
various edits with nico

>>>>

\TODO integrate the following text into the new intro

Recall that partial evaluators operate on multivariate functions $f$, whose
inputs are each labeled either {\em static} or {\em dynamic}. Once the static
inputs are provided, partial evaluation of $f$ produces a residual depending on
only the dynamic input. Mathematically, a partial evaluator is any $p$ such
that
\[
	\forall f,x. \exists f_x. \left[p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)\right]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to its
mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
specialized to $x$}.

\ur{  Based on this definition, it
  might be good and correct to argue that this is more or less an
  orthogonal problem.}

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before {\em any arguments} are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

### 2014-10-06 paper/semantics.tex
various edits with nico

Except for the staging features $\next$, $\prev$, and $\pause$, the language
features are essentially agnostic to stage. Variables in the context are
annotated with the stage at which they were introduced and can only be used at
the same stage.

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (future context).
The interrelation of these judgments can be summarized as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.
Because of the non-duplicating behavior, this sometimes produces explicit substitutions,
which are aggregated into a single {\em future context} that is also the output of evaluation.
\item Speculation sends a stage~\bbtwo\ term to a residual.
\item Reification sends a residual and a future context to another residual.  
\item Evaluation depends on speculation where stage~\bbone\ encapsulates stage~\bbtwo: $\next$. 
\item Speculation depends on evaluation where stage~\bbtwo\ encapsulates stage~\bbone: $\prev$ and $\pause$.
Reification is also used at these sites to handle the future context produced by evaluation.
\end{itemize}

Partial values are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\begin{lstlisting} 
1`(next {`2`yhat`1`},7)`
\end{lstlisting}
is a partial value.
Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\begin{lstlisting} 
2`let zhat = yhat*yhat*7 in zhat`
\end{lstlisting}
is a residual,
These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
They are different in that ``partial value" is a classification of stage~\bbone\ terms,
while ``residual" is a classification of stage-\bbtwo\ terms.

@@

At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.

Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
environment which maps that variable to the residual ($\hat y \mapsto q$).

@@

These dynamics allow us to define a partial evaluator.
Recall that partial evaluators operate on multivariate functions $f$, whose
inputs are each labeled either {\em static} or {\em dynamic}. Once the static
inputs are provided, partial evaluation of $f$ produces a residual depending on
only the dynamic input. Mathematically, a partial evaluator is any $p$ such
that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to its
mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
specialized to $x$}.

@@

We define a partial evaluator for \lang\ by identifying static with stage
\bbone\ and dynamic with stage \bbtwo.  Specifically, we encode $f$ as a \lang\
expression with a function type of the form $A\to\fut(B\to C)$.%
%\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
%the following higher-order function which makes the adjustment:
%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%  fn (p : int) => 
%    next{
%      fn (b : int) => 
%        prev{f (next {b}, p)}
%    }
%\end{lstlisting}}
%
Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.

>>>>

Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
language features (i.e., the introductory and eliminatory forms for product,
sum, and function types) preserve the stage of their subterms, and work at both
stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
\lang\ would simply consist of two non-interacting copies of a standard lambda
calculus. Variables in the context are annotated with the stage at which they
were introduced and can only be used at the same stage.

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (future context).
These judgments are related as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.
In order to avoid duplicating subterms, evaluation collects explicit
substitutions into a single {\em future context} that is also the output of
evaluation.
\item Speculation sends a stage~\bbtwo\ term to a residual.
\item Reification sends a residual and a future context to another residual.  
\item Evaluation depends on speculation for $\next$ terms, which have
stage~\bbtwo\ subterms.
\item Speculation depends on evaluation for $\prev$ and $\pause$, which have
stage~\bbone\ subterms. The future context produced by evaluation is also
reified here.
\end{itemize}

\emph{Partial values} are stage~\bbone\ terms that have been fully
evaluated, but which may contain stage~\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\begin{lstlisting} 
1`(next {`2`yhat`1`},7)`
\end{lstlisting}
is a partial value.
\emph{Residuals} ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\begin{lstlisting} 
2`let zhat = yhat*yhat*7 in zhat`
\end{lstlisting}
is a residual.

Note that partial values and residuals are both terms for which stage \bbone\
computation has completed; however, partial values are stage~\bbone\ terms,
while residuals are stage~\bbtwo\ terms.

\ur{I would recommend using ``1-value'' and ``2-value'' instead of
  partial values and residuals.  The former are more accurate, more
  direct, more succinct, and does not lead to confusions with other
  residuals.}

\ur{Many incorrectly hyphenated ``partial values'' in this text.}

@@

At $\next~e$, in contrast, $\redonesym$ speculates into $e$ for stage~\bbone\ 
subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.
Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
environment which maps that variable to the residual ($\hat y \mapsto q$).

@@

@@

These dynamics allow us to define a partial evaluator for \lang\ by identifying
static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
encode $f$ as a \lang\ expression with a function type of the form
$A\to\fut(B\to C)$.%
%\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
%the following higher-order function which makes the adjustment:
%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%  fn (p : int) => 
%    next{
%      fn (b : int) => 
%        prev{f (next {b}, p)}
%    }
%\end{lstlisting}}
%
Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.

### 2014-10-06 paper/splitting.tex
various edits with nico

\section{Splitting Algorithm}
\label{sec:splitting}
\ur{First para should probably be somewhere else concerning problem
  definition and relationship to PE.  Based on this definition, it
  might be good and correct to argue that this is more or less an
  orthogonal problem.}

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before {\em any arguments} are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

@@

We now consider the problem of splitting stage-\bbone\ partial values.
This serves as a simple didactic stepping-stone to the full formulation, 
and we will see later that it is a necessary component of the correctness theorem.

>>>>

\section{Splitting Algorithm}
\label{sec:splitting}

@@

We now consider the problem of splitting stage~\bbone\ partial values.
This serves as a simple didactic stepping-stone to the full formulation, 
and we will see later that it is a necessary component of the correctness theorem.
